Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-02-02 manu_draft.Rmd
methods revision from JH done

MOdel SYD adds an average within-year cyclic pattern as a separate spline:

@@

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

@@

The estimated seasonal average was calculated from the vector of daily values for each of the $n$ days in the season of interest with $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.  Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$. @Murphy19 provided a similar approach for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series.  For example, the above estimates were used for a test of the null hypothesis that `y` is the same at two times. To do so, $\bX$ would have two rows for each time to compare.  The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  This difference was used to determine if the two estimates in time were significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates. An alternative interpretation of this test is a quantification of differences between two points in time as affected by the magnitude of the difference and range of likely values for the magnitude of change.

@@

As a final analysis, the chosen sequence of years for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).  Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year (e.g., an estimate for 1990 for a six year window assesses the trend from 1987 to 1993).  Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

@@

Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)). All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at south Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.  The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).

@@

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year. However, the change from 2000 to 2010 was a significant increase of 64%, followed by a significant decrease of 35% from 2010 to 2019.  Although these trends were insignificant, the results are based on a priori decisions of which years to compare.

An assessment of the annual differences indicates if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences. Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but using seasonal averages from GAM results. Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to June.  Similarly, rows g-i show the seasonal averages and trend estimates from July to December.  The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.58 ug/L chlorophyll change per year).  Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.58)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.28). As a result, an assessment of trend changes by season provides a more precise assessment of trends over time and accounts for full uncertainty in estimates across model results.

Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although not all trends were significant and varied by location (Figure \@ref(fig:trndmap)a).  Comparing years 1990 and 2000, stations 21, 24, 27, and 30 had increasing significant trends.  The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.  Comparing 2000 and 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).  Comparing 2010 and 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.  The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to June and July to December (Figure \@ref(fig:trndmap)b).  For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at stations 30 and 32 from 2010 to 2019. For the July to December period, significant increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

A final analysis evaluated a moving window comparison of seasonal trends to provide additional context of when significant changes are occurring at each station (Figure \@ref(fig:winchg)).  Results were obtained using a centered five year window to estimate the slope and significance of a seasonal trend of GAM results from January to June and July to December for each five-year period at each station (note that estimates at the far left and far right of each time series cannot be calculated).  As before, larger trends were generally observed with the more southern stations.  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until recent years when trends were relatively stable.  Most stations showed a shift around 2007 when trends switched from significantly increasing to significantly decreasing, particularly in the January to June period.

@@

Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.  As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.  Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.  Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.

These distinctions are demonstrated in Figure \@ref(fig:trndcmpex) for two stations and two time periods, where the significance of trends are compared based on regression through seasonal averages (Figure \@ref(fig:trndcmpex) row 1), regression through GAM estimates (row 2), and mixed-meta regression through GAM estimates (row 3).  The stations and time periods were selected to demonstrate examples where different conclusions can be obtained depending on the methods and characteristics of the data.  For example, Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.  Only the mixed-meta regression results are significant in this example.  The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a significant trend.  Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.  Unlike the first example, the top two figures show significant trend estimates, whereas the bottom plot for the mixed-meta regression analysis does not show a significant trend because of added uncertainty in the averages provided by the GAMs.  In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.

Figure \@ref(fig:trndcmp) expands the results in Figure \@ref(fig:trndcmpex) by applying the method comparisons to all stations. As before, the different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade.  The slope estimates from the linear models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.  Slope estimates from the linear models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the linear models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

>>>>

Model SYD adds an average within-year cyclic pattern as a separate spline:

@@

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

@@

The estimated seasonal average was calculated from the vector of daily values for each of the $n$ days in the season of interest with $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.  Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$. @Murphy19 provided a similar approach for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series.

@@

### Trend comparisons

The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019. As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).  Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.  The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.  We chose a centered five-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year. Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

Finally, trend results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.  As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.  Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.  Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.  Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averaged. The comparisons were demonstrated with select examples where differences were pronounced and then applied to all stations.

@@

Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)). All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at South Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.  The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).

@@

To demonstrate the differences in results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)). All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.  The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}-Chla slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.  Chlorophyll also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).  Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10}-Chla slope -0.02, -0.04-0 95\% confidence interval).  Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.

Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.  Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.  
Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chlorophyll decreased for at stations 30 and 32 from 2010 to 2019. For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

@@

Results from a five-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).  The largest trends occurred at southern stations where increases occurred in both halves of the year (i.e., January to June, July to December).  Decreases were observed later in the record, until recent years when trends were  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until chlorophyll was relatively stable across all stations and seasons. A change from increasing to decreasing chlorophyll occurred at most stations around 2007, particularly for January to June.

Results showing trend estimates from mixed-meta regression on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2). Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.  Only the mixed-meta regression results shows a trend in this example (based on $p < 0.05$).  The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a trend.  Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.  Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend because of added uncertainty in the averages provided by the GAMs.  In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.  Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
